---
title: "afmr -- the beginning"
author: "Will Chase"
date: "November 8, 2018"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    theme: flatly

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Analysis of AFM images in R
This is a preview of the functionality of the afmr package I am currently developing. This package focuses on visualization and image analysis for atomic force microscopy (AFM) data. This tutorial utilizes my source code, but I have collected these functions as a package that is currently available on github (https://github.com/will-r-chase/afmr), but does not have all functions added yet. This document shows AFM data from the plant cell wall (collected by Will Chase). The plant cell wall is a fibrillar network, hence the images show the mesh-like texture of cellulose microfibrils that form this network. The code used in this document is hidden, but you can press the "code" buttons in the upper right to view the code for each section.

## Why?

1. Allows for batch processing
2. Allows for greater control over output
3. Allows for integration with other R tools

## Currently available: 

1. Read bruker scan files, compute matrices of any channel, extract parameters from header or add custom parameters
2. Visualize 2D maps and apply basic image analysis tools
3. Visualize 3D maps 

## To do:
1. Add more advanced visualization: 3D overlays and color schemes
2. Add common image transformations: plane fit, flatten, impute missing data, etc
3. Add support for interactive analysis (measurement, line drawing + histogram, etc)


# Test driving

## Setup

First load libraries and read functions
```{r libraries, message = FALSE, warning = FALSE}
#load packages (later afmr)
library(tidyverse)
library(rayshader)
library(EBImage)
library(assertthat)
```
```{r}
#extract parameters from the header
#parameter must be character vector that exactly matches the description in the header ie. "Tip Radius"
#if no match is found in the header, the parameter value is returned directly
#this allows users to store values that may not be in the header but are associated with the experiment
#such as "cell number" "treatment" etc.
afm_extract_parameter <- function(header = NULL, parameter = NULL){
  parameter_regex <- paste0("\\\\", parameter, ":")
  parameter_text <- header[str_detect(header, parameter_regex)][1]
  if (is.na(parameter_text)) {
    warning(paste(parameter, "was not found in the header, returning", parameter, "directly"))
    return(parameter)
  } else {
  tryCatch({readr::parse_number(parameter_text)}, 
           warning = function(w){str_extract(parameter_text, pattern = '(?<=: )(.*)(?=")')})
  }
}

#reads the scan data into a dataframe
afm_extract_scan_points <- function(data){
  assert_that(is.character(data))
  
  afm_df <- data %>%
    str_squish() %>%
    as.tibble() 
  
  names <- unlist(strsplit(x = as.character(afm_df[1, ]), split = " "))
  assert_that(is.character(names))
  
  afm_df %>%
    slice(-1) %>%
    {
      tryCatch(
        {tidyr::separate(., col = value, into = names, sep = " ", convert = TRUE)},
        warning = function(w) {stop("The rows of your scan data had an unequal 
                                    number of columns. Check that all lines of 
                                    your scan data are complete.")}
      )    
    }
}

#formats a data channel as a matrix for visualization
#provide channel argument as character vector that matches the colname in scan_points exactly
afm_matrix <- function(data, samps_line, afm_lines, channel = NULL){
  assert_that(!is.null(channel))
  
  data %>%
    pull(channel) %>%
    matrix(., ncol = samps_line, nrow = afm_lines, byrow = TRUE)
}

#takes required inputs (scan size, scan points, samps per line, lines)
#also takes optional params and maps (image matrices)
#formats as afm_scan object
afm_scan <- function(scan_points = NULL, scanSize = NULL, sampsPerLine = NULL, afmLines = NULL, maps = list(), optional_params = list()){
  assert_that(!is.null(scan_points), !is.null(scanSize), !is.null(sampsPerLine), !is.null(afmLines))
  assert_that(is.numeric(scanSize), is.numeric(sampsPerLine), is.numeric(afmLines))
  
  if(is.tibble(scan_points) == FALSE & is.data.frame(scan_points) == FALSE) {
    stop("scan_points must be a tibble or dataframe")
  }
  
  data <- list(
    params = list(scan_size = scanSize, samps_per_line = sampsPerLine, afm_lines = afmLines),
    scan_data = scan_points,
    opt_params = optional_params,
    maps = maps
  )
  
  class(data) <- "afm_scan"
  return(data)
} 

#reads a bruker text file that has a header and scan data from any number of channels
afm_read_bruker <- function(file = NULL, maps = "all", opt_params = list(), scan_size = "Scan Size", samps_per_line = "Samps/line", afm_lines = "Lines"){
  assert_that(!is.null(file))
  
  afm_text <- read_lines(file)
  afm_header <- afm_text[str_detect(afm_text, "\\\\")]
  afm_data <- afm_text[((2*length(afm_header))+1):length(afm_text)]
  
  assert_that(length(afm_header) > 0, msg = "No header found, your file must have a header")
  assert_that(length(afm_data) > 0, msg = "No scan data found, your file must have scan data")
  
  scanSize <- afm_extract_parameter(afm_header, scan_size)
  assert_that(is.numeric(scanSize), msg = "scan_size is not numeric, 
                                          check that it is present in 
                                          the header and that you correctly 
                                          specified the name of the parameter")
  sampsPerLine <- afm_extract_parameter(afm_header, samps_per_line)
  assert_that(is.numeric(sampsPerLine), msg = "samps_per_line is not numeric, 
                                          check that it is present in 
                                          the header and that you correctly 
                                          specified the name of the parameter")
  afmLines <- afm_extract_parameter(afm_header, afm_lines)
  assert_that(is.numeric(afmLines), msg = "afm_lines is not numeric, 
                                          check that it is present in 
                                          the header and that you correctly 
                                          specified the name of the parameter")
  scan_points <- afm_extract_scan_points(afm_data)
  
  if(maps == "all"){
    all_channels <- colnames(scan_points)
    map_names <- paste(str_extract(all_channels, pattern = ".+?(?=\\()"), "matrix", sep = "_")
    afm_maps <- map(.x = all_channels, ~afm_matrix(scan_points, sampsPerLine, afmLines, channel = .x))
    names(afm_maps) <- map_names
  } else{
    all_channels <- colnames(scan_points)
    select_channels <- maps[maps%in%all_channels]
    map_names <- paste(str_extract(select_channels, pattern = ".+?(?=\\()"), "matrix", sep = "_")
    afm_maps <- map(.x = select_channels, ~afm_matrix(scan_points, sampsPerLine, afmLines, channel = .x))
    names(afm_maps) <- map_names
  }
  
  optional_params <- purrr::map(opt_params, ~afm_extract_parameter(afm_header, .x))
  
  afm_scan(scan_points, scanSize, sampsPerLine, afmLines, maps = afm_maps, optional_params)
}
```

Next we can read in some data using the `afm_read_bruker()` function. I'll load a 500nm image and a 2um image.
```{r}
half_um_scan <- afm_read_bruker("500nm_1.txt", maps = c("Height_Sensor(nm)", "Peak_Force_Error(nN)"), opt_params = list(scan_rate = "Scan Rate", cap_dir = "Capture direction", num = 4, cell = "cell1"))
two_um_scan <- afm_read_bruker("2um_1.txt")
```

The structure of these files is a list with the various pieces organized for easy handling. 
```{r}
str(two_um_scan)
```

## Visualization

First we can analyze some 2D peakforce images:
```{r}
nanoscope_colors2 <- c("#000000", "#000000", "#0A0000", "#140000", "#1E0000", "#280000", "#320000", "#3C0000", "#460000", "#500000", "#5A0400", "#641500", "#6E2300", "#783400", "#824300", "#8C5100", "#966100", "#A07102", "#AA801B", "#B49037", "#BE9F52", "#C8AD6C", "#D2BD87", "#DCCCA3", "#E6DCBF", "#F0ECDB", "#FAFAF5", "#FFFFFF", "#FFFFFF")

nanoscope_palette <- colorRampPalette(nanoscope_colors2)
pf_matrix <- two_um_scan$maps$Peak_Force_Error_matrix
pf_scaled <- pf_matrix + abs(min(pf_matrix))
pf_image <- Image(pf_scaled)
pf_rot <- rotate(x = pf_image, -90)

pf_nanoscope <- colormap(pf_rot, nanoscope_palette(256))
pf_grey <- colormap(pf_rot, grey.colors(256))

display(pf_nanoscope, method = "raster")
display(pf_grey, method = "raster")
```

There's also some fancy stuff we can do like histogram equalization, gaussian blur, laplacian filtering

```{r}
eq_pf_nanoscope <- equalize(pf_nanoscope)
display(eq_pf_nanoscope, method = "raster")

fhi = matrix(1, nrow = 3, ncol = 3)
fhi[1, 1] = -7
img_fhi = filter2(pf_rot, fhi)
EBImage::display(img_fhi, method = "raster")
```

You can also do all sorts of normal things like cropping, rotating, false coloring, thresholding, making gifs, adding overlays, enhancing contrast/brightness

Of course, you have the raw data on hand, so statistical calculations are very easy. Here's some simple histograms for height and modulus channels. This is an area that can be expanded upon a lot. 

```{r}
graphics::hist(two_um_scan$scan_data$`Height_Sensor(nm)`, 20)
graphics::hist(two_um_scan$scan_data$`DMTModulus(MPa)`, 20)
```

The original reason I started this was to enable exportable 3D visualization. Here I show how to use the rayshader package to make a 3D representation of the 500nm height scan. This 3D object can be colored differently, saved as a spinning gif, embedded online as an interactive object, converted directly to 3D printing format, and more! I'm currently working on adding overlays (ie. modulus coloring with transparency) to the 3D object. 

```{r eval = FALSE}
height_matrix <- half_um_scan$maps$Height_Sensor_matrix

height_matrix %>%
  sphere_shade(texture = "desert",progbar = FALSE) %>%
  add_shadow(ray_shade(height_matrix,zscale=0.8,maxsearch = 300,progbar = FALSE),0.7) %>%
  plot_3d(height_matrix, zscale=0.8,fov=0,theta=0,zoom=0.9,phi=45, windowsize = c(800,800), solid = TRUE)
```

<video width=100% controls>
  <source src="out.mp4" type="video/mp4">
</video>
